---
title: "Forward Mode AD in R"
author: "Jonny Law"
date: "2019-08-05"
slug: forward-mode-automatic-differentiation-R
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hedgehog)
```

## Forward Mode Automatic Differentation

Automatic differentiation can be used to calculate the exact derivative of a function at a point using applications of the chain rule. Dual numbers provide a straightforward implementation in R using S3 generic methods. A dual number has a real component and a "dual" component which can be used to exacly calculate the expression and derivative at a specific value of $x$. Consider the quadratic form $f(x) = 5x^2 + 3x + 10$ with derivative $f^\prime(x) = 10x + 3$. The function and derivative can be evaulated at a value, say $x = 5$ using the dual number $5 + \varepsilon$, the dual component $\varepsilon$ is considered small such that $\varepsilon^2 = 0$ then calculating $f(5 + \varepsilon)$:

$$\begin{align}
f(5 + \varepsilon) &= 5(5 + \varepsilon)^2 + 3(5 + \varepsilon) + 10,\\
&= 5(25 + 10\varepsilon + \varepsilon^2) + 15 + 3\varepsilon + 10,\\
&= 5\varepsilon^2 + 53\varepsilon + 150.
\end{align}$$

Then the coefficient of $\varepsilon$ is the deriviative and the constant is the evaluation of the function, $f(5) = 150$ and $f^\prime(5) = 53$.

## S3 Objects

R has three systems for object oriented programming, S3, S4 and reference classes which can be learned about in the relevant chapter of [Advanced R](http://adv-r.had.co.nz/OO-essentials.html). Dual numbers can be implemented as an S3 class in R:

```{r}
dual <- function(real, eps) {
  if (!is.numeric(real)) stop("real must be numeric")
  structure(list(real = real, eps = eps), class = "dual")
} 
var <- function(x) {
  if (!is.numeric(x)) stop("x must be numeric")
  dual(x, 1)
}
const <- function(x) {
  if (!is.numeric(x)) stop("x must be numeric")
  dual(x, 0)
}
```

`var` represents a variable which we want to differentiate, whereas `const` represents a constant. 

Next, primitive functions can be defined in terms of `dual` numbers which simultaneously evaluate the function and the derivative:

```{r}
plus <- function(x, y) 
  dual(x$real + y$real, x$eps + y$eps)
minus <- function(x, y) 
  dual(x$real - y$real, x$eps - y$eps)
times <- function(x, y) 
  dual(x$real * y$real, x$eps * y$real + y$eps * x$real)
divide <- function(x, y) 
  dual(
      x$real / y$real,
      (x$eps * y$real - x$real * y$eps) / (y$real * y$real)
    )
```


Group generics can be used to implement the mathematics of dual numbers. Group generics included with base R include `Math` which includes special functions such such as `abs` and `sqrt` as well as trigonometric and hyperbolic functions. `Ops` which include the basic infix operations reqruired for arithmetic, `+`, `-`, `*`, `/` etc. For a full list of group generics associated with `Math` and `Ops` consult the R help by typing `?groupGeneric` in the R console. In order to implement a group generic for the S3 class `dual` we implement `Ops.dual`:

```{r}
Ops.dual <- function(x, y) {
  switch(
    .Generic,
    `+` = plus(x, y),
    `-` = minus(x, y),
    `*` = times(x, y),
    `/` = divide(x, y)
  )
}
```

`switch` is used to pattern match on the generic function being called within `Ops` by matching on `.Generic`. Implementing dual numbers in this way allows us to define a function using the in-built infix operators in a natural way. The function $f(x)$ can be defined in terms of dual numbers as

```{r}
f <- function(x) 
  const(5) * x * x + const(3) * x + const(10)
```

Then evaluated at $x = 5$ using the constructor `var` which initialises a dual with $\varepsilon = 1.0$.

```{r}
f(var(5))
```

The definition of `f` is cumbersome since we have to explicitly create the constants using the `const` constructor. The methods defined in `Ops.dual` can be extended to handle cases when a double is multiplied by a dual number to convert the double to a `const` and hence we can automatically differentiate any univariate function using forward mode automatic differentiation.

We can write a function which checks the arguments of `plus`, `minus` etc, then if the arguments aren't explicitly dual number variables using the function `var` then they are converted to a dual constant using `const`. This function checks each argument (of a generic function of two arguments `f`) in turn to determine if they are doubles then promotes them to constants.

```{r}
lift_function <- function(f) {
  function(x, y)
    if (is.double(x)) {
      f(const(x), y)
    } else if (is.double(y)) {
      f(x, const(y))
    } else {
      f(x, y)
    }
}
```

The ops can then be re-defined using the `lift_function`:

```{r}
Ops.dual <- function(x, y) {
  switch(
    .Generic,
    `+` = lift_function(plus)(x, y),
    `-` = lift_function(minus)(x, y),
    `*` = lift_function(times)(x, y),
    `/` = lift_function(divide)(x, y)
  )
}
```

Then `f` can be defined more naturally:

```{r}
f <- function(x) 
  5 * x * x + 3 * x + 10
```

And the derivative calculated:

```{r}
f(var(5))
```

## Testing using Hedgehog

[Hedgehog](https://github.com/hedgehogqa/r-hedgehog) is a package which utilises [testthat](https://testthat.r-lib.org) to implement property based testing in R. Property based testing can be used to check a wide range of inputs to a function and determine if the code outputs the expected value. In standard unit testing the state before the test is defined by programmer and typically does not change - if we were to consider a test for the derivative of the quadratic function defined above then we might write a test which evaluates the function at $x = 5$. This verifies we are correct for $x = 5$, but what about $x = 0$ or another value. With property based testing, we define a random generator for the input and the test checks hundreds of potential values for failure.

The input to this property based test is `a`, a number between $-100$ and $100$. The usual `testthat` syntax is then used to evaluate the gradient using forward mode AD and comparing it to the exact derivative calculated by hand. 

```{r}
test_that("Derivative of 5x^2 + 3x + 10",
          forall(list(a = gen.c(gen.element(
            -100:100
          ))),
          function(a)
            expect_equal(object = f(var(a))$eps, expected = 10 * a + 3)))
```