---
title: Harrier League Cross Country
author: Jonny Law
date: '2017-10-26'
slug: harrier-league-cross-country
categories:
  - R
  - Running
tags: []
---

```{r setup, include=FALSE}
library(tidyr); library(dplyr); library(readr); library(ggplot2); library(jonnylaw);
library(htmltab)
```

The [Harrier League](http://www.harrierleague.com) is a cross country
running league with seven fixtures across the North East of England in
the 2017/18 season across the winter months from September ’17 until
March ‘18.

The Harrier League is unique to other cross country fixtures because the
senior runners are divided up into slow, medium and fast packs. In the
senior men's race, the slow runners start first followed 2 minutes 30
seconds later by the medium pack runners, then a further 2 minutes 30
seconds by the fast pack runners.

In order to progress to the fast pack, runners must first run in the
slow pack and finish in the top 10% of finishers, this entitles them to
run from the medium pack. If they then finish in the top 10% from the
medium pack, then they can run from the fast pack.

Teams are split into three divisions and final team positions are
calculated within each division by ordering the first six runners (four
in the senior women) by race time. This is the total time elapsed from
when the slow pack starts to when the runner crosses the finish line.
Let's answer one interesting question, what would the results be from
the past weekends fixture without the handicaps, i.e. if everyone started
from scratch.

The code for this analysis was written using
[R](https://www.r-project.org/) and the
[tidyverse](https://github.com/tidyverse/). The full code can be found
on
[Github](https://gist.github.com/jonnylaw/e882d245188a4b7c38425cc5e0eddbd6).
Let me know if you’d like me to answer other questions about the harrier
league.

# Senior Womens

We will look at the results from the previous race at Druridge Bay.
In the senior women's race there are four counters, any team with less
than four runners will be removed from the final results. Now, we can
calculate the position of each runner within each division by the actual
running time of each runner.

First we need to download the relevant information, since the results are available as an HTML table, the package [htmltab](https://cran.r-project.org/web/packages/htmltab/index.html) can be used to select the table containing the individual results

```{r import_info, message=FALSE, warning=FALSE}
womens_results_raw = htmltab::htmltab(doc = "http://www.harrierleague.com/results/2017-18/druridge/SenF.htm")
```

These results are then cleaned up, removing any guest runners and extracting the division of each runner from the club field. 

```{r process_results}
clean_results = function(raw_results) {
  raw_results %>%
    as_data_frame() %>%
    rename(race_time = `Race Time`, actual_time = `Actual Time`) %>%
    filter(Cat != "guest") %>%
    mutate(division = as.numeric(stringr::str_sub(Club, start = 2, end = 2)), 
           club = stringr::str_sub(Club, start = 5)) %>%
    select(Name, actual_time, division, club) %>%
    mutate(actual_seconds = as.numeric(lubridate::ms(actual_time)))
} 

womens_results = clean_results(womens_results_raw)
```

Now we write a function which calculates the results for one division. This can be re-used for the men's results by accepting the number of counters for each team as an argument to the function (4 women to count, 6 men to count). We remove any clubs who fielded incomplete teams and select the top counters ordered by the position supplied in the `raw_results` dataframe. We then calculate the total points for each time, by summing the positions (lower is better) and concatenate the names of the counters into a single string.

```{r}
concat = function(strings) {
  Reduce(f = function(string_1, string_2) paste(string_1, string_2, sep = ", "), x = strings)
}

get_results_one_division = function(raw_results, counters = 4) {
  incomplete_teams = raw_results %>%
    count(club) %>%
    filter(n < counters)
  
  raw_results %>%
    anti_join(incomplete_teams, by = "club") %>%
    group_by(club) %>%
    top_n(n = counters, wt = desc(pos)) %>%
    group_by(club) %>%
    summarise(total_points = sum(pos), total_counters = n(), 
              counters = concat(paste0(Name, " (", pos, ")"))) %>%
    mutate(position = min_rank(total_points))
}
```

Now, we are ready to calculate the results given positions. But instead of using the race time (actual time + handicap time), we use the actual time to determine individual positions within each of the three divisions. We then use the previously defined function `get_results_one_division()` to calculate the final score for each team in each division: 

```{r, results='asis'}
actual_time_results = function(results, counters) {
  results %>%
    group_by(division) %>%
    mutate(pos = min_rank(actual_seconds)) %>%
    do(get_results_one_division(., counters = counters)) %>%
    arrange(division, position)
} 
```

Here are the results for the women's division one, two and three:

```{r final_results_division_one, echo=FALSE, results='asis'}
actual_time_results(womens_results, counters = 4) %>% 
  filter(division == 1) %>%
  ungroup() %>%
  select(position, club, total_points, counters) %>%
  knitr::kable(align = "c", caption = "Senior Women's Division One")
```

```{r division_two_women, echo=FALSE, results='asis'}
actual_time_results(womens_results, counters = 4) %>% 
  filter(division == 2) %>%
  ungroup() %>%
  select(position, club, total_points, counters) %>%
  knitr::kable(align = "c", caption = "Senior Women's Division Two")
```

```{r division_three_women, echo=FALSE, results='asis'}
actual_time_results(womens_results, counters = 4) %>% 
  filter(division == 3) %>%
  ungroup() %>%
  select(position, club, total_points, counters) %>%
  knitr::kable(align = "c", caption = "Senior Women's Division Three")
```

## Senior Men

In the senior men's race there are six counters, any team with less than
six runners will be removed from the final results. Now, we can
calculate the position of each runner within each division by the actual
running time of each runner.

```{r read_means_results, message=FALSE, warning=FALSE}
mens_results_raw = htmltab(doc = "http://www.harrierleague.com/results/2017-18/druridge/SenM.htm")
mens_results = clean_results(mens_results_raw)
```


```{r division_one_men, echo=FALSE, results='asis'}
actual_time_results(results = mens_results, counters = 6) %>% 
  filter(division == 1) %>%
  ungroup() %>%
  select(position, club, total_points, counters) %>%
  knitr::kable(align = "c", caption = "Senior Men's Division One")
```

```{r division_two_men, echo=FALSE, results='asis'}
actual_time_results(results = mens_results, counters = 6) %>% 
  filter(division == 2) %>%
  ungroup() %>%
  select(position, club, total_points, counters) %>%
  knitr::kable(align = "c", caption = "Senior Men's Division Two")
```

```{r division_three_men, echo=FALSE, results='asis'}
actual_time_results(results = mens_results, counters = 6) %>% 
  filter(division == 3) %>%
  ungroup() %>%
  select(position, club, total_points, counters) %>%
  knitr::kable(align = "c", caption = "Senior Men's Division Three")
```