---
title: "The Kalman Filter in Scala"
author: "Jonny Law"
date: '2016-12-12T12:13:14-05:00'
categories: scala
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyr); library(dplyr); library(readr); library(ggplot2); library(jonnylaw);
```

A Dynamic Linear Model (DLM) is a special type of state space model, where the state and observation equations are Normally distributed and linear. A general DLM can be written as follows:

$$\begin{aligned} y_t &= F_t x_t + \nu_t, \qquad \nu_t \sim \mathcal{N}(0, V_t) \\
x_t &= G_tx_{t-1} + \omega_t \qquad \omega_t \sim \mathcal{N}(0, W_t), \end{aligned}$$

$y_t$ represents the observation of the process at time $t$, $x_t$ is the value of the unobserved state at time $t$. The observation error $\nu_t$ and the system error $\omega_t$ are independent and identically distributed (iid) Normal random variables. $F_t$ is the observation matrix which transforms the state space to the observation, $G_t$ is the state transition matrix.

## Forward Simulating from the DLM

A first order polynomial DLM with constant $V$ and $W$, and $F_t = 1$, $G_t = 1$. can be simulated in scala as follows:

```scala
import breeze.stats.distributions.Gaussian

case class Data(time: Time, observation: Observation, state: Option[State])
case class Parameters(v: Double, w: Double, m0: Double, c0: Double)

def simulate(p: Parameters): Stream[Data] = {
val stateSpace = Stream.iterate(Gaussian(p.m0, sqrt(p.c0)).draw)(x => Gaussian(x, sqrt(p.w)).draw)
  stateSpace.zipWithIndex map { case (x, t) =>
    Data(t, x + Gaussian(0, sqrt(p.v)).draw, Some(x)) 
  }
}

val p = Parameters(3.0, 0.5, 0.0, 10.0)
// simulate 16 different realisations of 100 observations, representing 16 stations
val data = (1 to 16) map (id => (id, simulate(p).take(100).toVector))
```

We use the built in streaming library's `iterate` function to specify the evolution of the latent state. The initial state $x_0$ is a sample drawn from a normal distribution with mean $m_0$ and variance $C_0$, $x_t \sim \mathcal{N}(x_0 ; m_0, C_0)$. The [Breeze numerical computing library](https://github.com/scalanlp/breeze/) provides many statistical and mathematical functions  is used. Subsequent states are generated by adding $\mathcal{N}(0, W)$ to the previous state:

$$x_t = x_{t-1} + \omega, \qquad \omega \sim \mathcal{N}(0, V)$$

We then construct a `Stream` of `Data` objects. The `Data` object has a timestamped observation and an optional state space. We construct the observation at time $t$ by simply adding the observation noise to the state space $y_t \sim \mathcal{N}(y_t | x_t, V)$. The state is optional because we can't observe the state of real data, only simulated data will have a known state.

A graph of the data from four "stations", produced using [ggplot2](http://ggplot2.org) in R is shown in the figure below.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
data = read_csv(here::here("notebooks/data/firstOrderDlm.csv"), col_names = c("stationId", "time", "observation", "state"))

## clean the data using dplyr and plot using ggplot2
data %>%
  gather(key = "key", value = "value", -time, -stationId) %>%
  filter(stationId %in% 1:4) %>%
  ggplot(aes(x = time, y = value, colour = key)) + 
  geom_line() +
  facet_wrap(~stationId, scales = "free")
```

### An Aside on Referential Transparency

Note that the function `simulate` is not referentially transparent, meaning the function will return a different `Stream` of data each time we run it. Referential transparency is important in functional programming, to allow us to easily reason about complex programs. The Breeze library implements another object for stateful random number generation, the `Process` object. The `MarkovChain` object can be used to construct a process without drawing explicitly from the distribution until we run the program. Firstly define a single step of the random walk:

```scala
import breeze.stats.distributions._

def step_rw(p: Parameters): Double => Rand[Double] = 
  x => Gaussian(x, p.w)
```

If `step_rw` is supplied with a set of `Parameters`, it returns a function from the current state, which is assumed to be materialised, to the next state, which is a `Rand[Double]`. The actual random number isn't generated until we sample from distribution represented by `Rand`. Next, we can construct a `MarkovChain` using the transition kernel `stepRw`:

```scala
val random_walk: Process[Double] = MarkovChain(0.0)(step_rw(p))
random_walk.
  steps.
  take(100)
```

## The Kalman Filter

The Kalman Filter can be used to determine the posterior distribution of the state space given the current observation and the $p(x_t|D_{t})$, where $D_t = \{Y_t, D_{t-1}\}$ and $D_0$ is the initial information, comprising of the parameters mode parameters $W_0$ and $V_0$ and the initial state $x_0 \sim N(m_0, C_0)$. The full treatment of the Kalman Filter can be found in the excellent Bayesian Forecasting for Dynamic Models by West and Harrison.

I will present the filtering equations for the simple model in this post. Suppose we start with the posterior distribution of $x_{t-1} \sim N(m_{t-1}, C_{t-1})$. The first thing we need to do is advance the state, the equation to advance the state is a simple Markov transition $x_t = x_{t-1} + \omega_t$. We simply add the system variance, the system variance is drawn from a Normal distribution with zero mean and variance $W$. The [sum of two Normal distributions is Normal](https://en.wikipedia.org/wiki/Sum_of_normally_distributed_random_variables) with the mean and variance added, hence the prior for $(x_t | D_{t-1}) \sim N(m_{t-1}, C_{t-1} + W)$.

Next we need to construct the observation, using the observation equation which is commonly called the one-step forecast for the series, $y_t = x_t + \nu_t$, since $\nu_t$ is Normally distributed with zero mean and variance $V$, the distribution of the one step forecast is, $(y_t|D_{t-1}) \sim N(m_{t-1}, C_{t-1} + W + V)$.

Now, we observe the true value of $y_t$ and are able to construct the posterior distribution of $(x_t | D_t) \sim N(m_t, C_t)$. $m_t = m_{t-1} + A_t e_t$ and $C_t = A_tV$. $A_t = \frac{C_{t-1} + W}{ C_{t-1} + W + V}$ is known as the regression coefficient, and $e_t = Y_t - m_{t-1}$. This result can be shown using properties of the multivariate normal distribution, and is presented in full in Bayesian Forecasting for Dynamic Models by West and Harrison.

We now have the equations required to program up the Kalman Filter using Scala.

```scala
case class FilterState(data: Data, p: Parameters)

def filter(p: Parameters): (FilterState, Data) => FilterState = (s, d) => {
  val r = s.p.c0 + p.w
  val q = r + p.v
  val e = d.observation - s.p.m0

  // kalman gain
  val k = r / q
  val c1 = k * p.v
  
  // return the data with the expectation of the hidden state and the updated Parameters
  FilterOut(Data(d.time, d.observation, Some(m1)), Parameters(p.v, p.w, m1, c1))
} 
```

The function filter simply takes in `Parameters` and one observation, represented by `Data` and returns the updated parameters required for the next step of the filter. Now we need to write a function which filters a sequence of `Data`, and returns a sequence consisting of the latent states, which we can do using the function `scanLeft`.

A simplified function signature of `scanLeft` is given by: `scanLeft[A](l: List[A], z: A)(f: (A, A) => A): List[A]`. A list, `l` with elements of type `A` and an initial value, `z` also of type `A` if passed to a `Function2` and accumulated into another list with elements of type `A`. The function `f` is applied to each element of the list pairwise, starting the the head of the lift and the zero element, `z`. Consider calculating the `sum` of a list of numbers:

```scala
val numbers = List(1,2,3,4,5)
def sum(a: Int, b: Int): Int = a + b

numbers.scanLeft(0)(sum)
// List(0, 1, 3, 6, 10, 15)
```

The first calculation is (0 + 1) = 1, which is then used as the first argument in the pairwise `sum` function, then the second calculation is (1 + 2) = 3, the result of which is used again in the next application of `sum`. Each intermediate step of the calculation is retained and appended to a list to be output when the list of number is exhausted, so we end up with a cumulative sum, `List(0, 1, 3, 6, 10, 15)`. We can use `scanLeft` and the `Function2`, `filter` to calculate and retain the latent states in a DLM:

```scala
def filterSeries(data: Seq[Data], initParams: Parameters): Seq[FilterOut] = {
  val initFilter = FilterState(data.head, params, 0.0) // initialise the filter

  data.
    scanLeft(initFilter)(filter(initParams)).
    drop(1)
}
```

Now, we can apply the filter to all the stations simultaneously:

```scala
data.
  groupBy{ case (id, _) => id }. //groups by id
  map{ case (id, idAndData) =>
  (id, idAndData map (x => x._2)) }. // changes into (id, data) pairs
  map{ case (id, data) =>
  (id, filterSeries(data.sortBy(_.time), p)) } // apply the filter to the sorted data
```

We can now plot the results of the filtering using R and ggplot2, overlaid with 95% prediction intervals.

```{r}
filtered = read_csv(here::here("notebooks/data/filteredDlm.csv"), c("stationId", "time", "observation", "stateMean", "m", "c"))

## calculate upper and lower 95% bounds of the state estimate
filtered %>%
  select(-observation) %>%
  inner_join(data, by = c("time", "stationId")) %>%
  filter(stationId %in% 1:4) %>%
  mutate(upper = qnorm(p = 0.975, mean = m, sd = sqrt(c)), 
         lower = qnorm(p = 0.025, mean = m, sd = sqrt(c))) %>%
  select(-c, -m, -observation) %>%
  gather(key, value, -time, -stationId, -upper, -lower) %>%
  ggplot(aes(x = time, y = value, linetype = key)) + 
  geom_line() +
  geom_ribbon(aes(x = time, ymin = lower, ymax = upper), alpha = 0.3) +
  facet_wrap(~stationId, scales = "free")
```

The full code is available in a notebook file and an [ammonite](https://ammonite.io/) script in the [GitHub Repo](https://github.com/jonnylaw/blog/notebooks) associated with this blog.