---
title: Seasonal DLM
author: Jonny Law
date: "2016-12-13T12:13:14-05:00"
---

```{r setup, include=FALSE}
library(tidyr); library(dplyr); library(readr); library(ggplot2); library(jonnylaw);
library(scales)
theme_set(theme_minimal())
```

## The Seasonal DLM

I introduced the class of state space models called DLMs in a previous post covering the [Kalman Filter](KalmanFilter/). The seasonal DLM is similar to the first order DLM, however it incorporates a deterministic transformation to the state, in order to capture cyclic trends. Remember a general DLM can be written as:

$$\begin{align}
y_t &= F_t x_t + \nu_t, \qquad \mathcal{N}(0, V_t), \\
x_t &= G_t x_{t-1} + \omega_t, \quad \mathcal{N}(0, W_t).
\end{align}$$

In the fourier-form seasonal DLM, the state is transformed by a block diagonal matrix, $G_t$, containing rotation matrices. The rotation matrix is given by:

$$R = \begin{pmatrix}
\cos(\omega) & -\sin(\omega) \\
\sin(\omega) & \cos(\omega)
\end{pmatrix},$$
where $\omega$ is the frequency of the seasonality. In practice, it is easier to specify the period of the seasonality, $T$, which is related to the frequency: $\omega = 2 \pi/T$. This means if we have data measured at hourly intervals and we believe the process has a daily cycle, then we will set $T = 24$.

In order to model higher harmonics of the seasonality, we combine rotation matrices together into a block diagonal matrix, to form the $G$ matrix, for instance with 3 harmonics the system matrix is:

$$G = \begin{pmatrix}
\cos(\omega) & -\sin(\omega) & 0 & 0 & 0 & 0\\
\sin(\omega) & \cos(\omega) & 0 & 0 & 0 & 0 \\
0 & 0 & \cos(2\omega) & -\sin(2\omega) & 0& 0 \\
0 & 0 & \sin(2\omega) & \cos(2\omega) & 0 & 0 \\
0 & 0 & 0 & 0 & \cos(3\omega) & -\sin(3\omega) \\
0 & 0 & 0 & 0 & \sin(3\omega) & \cos(3\omega) \\
\end{pmatrix}.$$

In this case, the latent state, $x_t$ is six-dimensional. This means system evolution variance-covariance matrix, $W_t$ is a six by six matrix.

First we should make a `case class` in Scala representing a DLM. A `case class` in Scala is a class with a default `apply` method used to construct instances of the class. The `case class` also has getter methods which can be used to access the values of the class. The `case class` will be for a DLM with constant $F$ and $G$ matrices. 

```scala
import breeze.linalg._

case class Dlm(f: DenseMatrix[Double], g: DenseMatrix[Double])
```

Note that, for univariate observations $F$ will be a row vector, the recommended way to specify row vectors using [Breeze](https://github.com/scalanlp/breeze/) is to use a matrix with a single row. The DLM also has associated parameters which we will assume are constant and write as a `case class`.

```scala
case class Parameters(v: Double, w: DenseMatrix[Double], m0: DenseVector[Double], c0: DenseMatrix[Double])
```

Then we can add a method to simulate forward given a value of the `Parameters`, firstly we will write a single step of the simulation:

```scala
case class Data(time: Double, observation: Double, state: DenseVector[Double])

def simStep(p: Parameters, model: Dlm): Data => Rand[Data] = d => {
    for {
      w <- MultivariateGaussian(DenseVector.zeros(p.w.cols), p.w)
      x1 = model.g * d.state + w
      v <- Gaussian(0, p.v)
      y = model.f.toDenseVector dot x1 + v
    } yield Data(d.time + 1.0, y, x1)
  }
```

This function contains a for comprehension, the `<-` symbol represents either a `map` or a `flatMap`. Since the `w` and `v` are `Rand[Double]` values, we need to access the value inside of the `Rand` and perform a function on it, the for comprehension desugars to:

```scala
MultivariateGaussian(DenseVector.zeros(p.w.cols), p.w).
  flatMap(w => {
    val x1 = model.g * d.state + w
    Gaussian(0, p.v).map(v => {
      val y = model.f.toDenseVector dot x1 + v
      Data(d.time + 1.0, y, x1)
    })}
    )
```

The desugared chain of `flatMap` and `map` evaluates to the same result, but the for-comprehension is more readable. This syntax provides a clean and elegent way to work within the context of a Monad, `Rand` is an example of a Monad representing a distribution.

Now, we can use the `MarkovChain` breeze object to simulate the full series:

```scala
def simMarkov(p: Parameters, model: Dlm): Rand[Process[Data]] = {
  for {
    x0 <- MultivariateGaussian(p.m0, p.c0)
    y0 <- Gaussian(model.f.toDenseVector dot x0, p.v)
    init = Data(0.0, y0, x0)
  } yield MarkovChain(init)(simStep(p, model))
}
```

Note that the Markov chain is a `Rand[Process[Double]]`, this is not the most elegant and I'm open to suggestions on a better way to formulate this. Once we have defined a Markov chain, we must draw from it which returns a `Process[Double]`, the initial draw is to sample from the initial state distribution $\mathcal{N}(m_0, C_0)$. Next, we can sample from the `Process` and plot it. A simulation from the seasonal DLM, with parameters $V = 3$ and $W = I_6$, is given below:

```{r, echo=FALSE, message=FALSE}
sims = read_csv(here::here("notebooks/data/SeasonalDlm.csv"), col_names = c("Time", "Observation", sapply(1:6, function(i) paste0("state_", i))))

sims %>%
  ggplot(aes(x = Time, y = Observation)) + 
  geom_line()
```